---
# title: Basics of **`ggplot2`**
format:
  revealjs:
    theme: FACE-IT_pres.scss
    self-contained: true
    transition: slide
editor: source
---

# Basics of **`ggplot2`** {background-color="#008980" style="text-align: center"}

*Robert Schlegel*

```{r plots-1-opts, echo=FALSE}
knitr::opts_chunk$set(
  warning = FALSE, 
  message = FALSE,
  echo = TRUE
)
```

## Problem

* Default R plotting (_Base R_) hasn't advanced much since the 90s
* Non-intuitive syntax, functions, and arguments
* Published figures do not look very professional

## Solution

* **`ggplot2`** uses the [grammar of graphics](https://link.springer.com/book/10.1007/0-387-28695-0)
* Integrated into **`tidyverse`** with intuitive functions and arguments
* Well developed support and extensions for professional figures

## Setup

We will need the following two packages for the examples in these slides.

```{r}
library(tidyverse) # Contains ggplot2

library(palmerpenguins) # Contains the dataset
```

## Basics

* One figure is made of one chunk of code
* Starts with `ggplot()`
* Each line is connected with a `+`
* Add shapes with `geom` functions
    * e.g. `geom_point()` adds points
* Plot skeleton created within `aes()`
* Arguments assigned like normal functions (e.g. `ggplot(data = penguins)`)

## Basic plot

```{r}
ggplot(data = penguins,
       aes(x = body_mass_g, y = bill_length_mm)) +
  geom_point(aes(colour = species))
```

## Focus on `aes()`

* Understand when to use `aes()`
* Columns from the data go inside `aes()`
    * e.g. `geom_point(aes(colour = species))`
* Static values go outside `aes()`
* Mistakes with `aes()` are common when learning **`ggplot2`** 
* Good starting point when looking for errors

## Inside `aes()` {auto-animate=true}

```{r}
ggplot(data = penguins,
       aes(x = body_mass_g, y = bill_length_mm)) +
  # 'island' is a column from 'penguins'
  geom_point(aes(colour = island))
```

## Outside `aes()` {auto-animate=true}

```{r}
ggplot(data = penguins,
       aes(x = body_mass_g, y = bill_length_mm)) +
  # Make all points red
  geom_point(colour = "red")
```

## Outside inside? {auto-animate=true}

```{r}
ggplot(data = penguins,
       aes(x = body_mass_g, y = bill_length_mm)) +
  # What is happening here?
  geom_point(aes(colour = "red"))
```

## Inside outside? {auto-animate=true}

```{r, error=TRUE}
ggplot(data = penguins,
       aes(x = body_mass_g, y = bill_length_mm)) +
  # Why does this cause an error?
  geom_point(colour = species)
```

## More than just colour {auto-animate=true}

```{r}
ggplot(data = penguins,
       aes(x = body_mass_g, y = bill_length_mm)) +
  # What else can we add?
  geom_point(aes(size = flipper_length_mm, shape = island))
```

## Add a geom {auto-animate=true}

```{r}
ggplot(data = penguins,
       # NB: Arguments passed to first aes() apply to all geoms
       aes(x = body_mass_g, y = bill_length_mm, colour = species)) +
  geom_point() +
  geom_smooth(method = "lm")
```

## Change labels {auto-animate=true}

```{r, echo=TRUE, warning=FALSE, message=FALSE}
ggplot(data = penguins,
       aes(x = body_mass_g, y = bill_length_mm, colour = species)) +
  geom_point() +
  geom_smooth(method = "lm") +
  # Change label text
  labs(x = "Body mass (g)", y = "Bill length (mm)", colour = "Species") +
  # Change legend position
  theme(legend.position = "bottom")
```

